---
title: "Introduction to causalreg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to causalreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `causalreg` package implements causal discovery via **Pearson risk
invariance** for generalized linear models (GLMs) and generalized additive
models (GAMs). Given a response and a set of candidate covariates, it
identifies which covariates are causal parents of the response within a
structural causal model.

The key idea: if a GLM or GAM is correctly specified with respect to the
true causal parents, the **Pearson risk** (average squared Pearson residual)
converges to 1. The package tests this property across subsets of covariates to
find the causal model.

## Causal Poisson regression

Consider a simple setting where `X1` causes `Y` (Poisson), and `X2` is a
downstream effect of `Y`:

```{r poisson-glm}
library(causalreg)

n <- 1000
set.seed(123)
X1 <- rnorm(n)
Y  <- rpois(n, exp(X1))
X2 <- log(Y + 1) + rnorm(n, 0, 0.3)
data <- data.frame(X1, X2, Y)
```

### Exhaustive search with chi-square test

For Poisson models, the chi-square test is fast and appropriate:

```{r poisson-all}
result <- cglm(Y ~ X1 + X2, "poisson", data, pval = "chi-square", search = "all")
result$model.opt
```

The method correctly identifies `Y ~ X1` as the causal model.

We can inspect the full results:

```{r poisson-all-details}
# All models considered
unlist(result$models)

# Their p-values (higher = more consistent with Pearson risk = 1)
result$pv

# Their BIC values
result$bic
```

Only the model `Y ~ X1` has a p-value above the significance threshold
(alpha = 0.05 by default), and it is selected.

### Stepwise search

For larger numbers of covariates, exhaustive search over all 2^p - 1
subsets becomes impractical. The stepwise search provides a faster
alternative:

```{r poisson-step}
result_step <- cglm(Y ~ X1 + X2, "poisson", data, pval = "chi-square", search = "stepwise")
result_step$model.opt

# Models visited during the search
unlist(result_step$models)
```

The stepwise search starts with an intercept-only model, then greedily adds
the variable that maximizes the Pearson risk p-value. After the forward
phase, it performs backward elimination based on BIC.

## Causal logistic regression

For binomial models, the chi-square approximation does not hold, so we use
the bootstrap test:

```{r binomial-glm}
n <- 2000
set.seed(123)
X1 <- rnorm(n)
Y  <- rbinom(n, 1, exp(X1) / (1 + exp(X1)))
flip <- rbinom(n, 1, 0.1)
X2 <- (1 - flip) * Y + rnorm(n, 0, 0.3)
data <- data.frame(X1, X2, Y)

set.seed(1)
result <- cglm(Y ~ X1 + X2, "binomial", data, pval = "bootstrap", search = "all")
result$model.opt
```

The bootstrap test correctly identifies the causal model despite the
binomial response.

## Causal GAMs for nonlinear relationships

When the relationship between covariates and response is nonlinear, use
`cgam()` with smooth terms:

```{r poisson-gam}
n <- 1000
set.seed(123)
X1 <- rnorm(n)
Y  <- rpois(n, exp(sin(X1)))
X2 <- log(Y + 1) + rnorm(n, 0, 0.5)
data <- data.frame(X1, X2, Y)

result <- cgam(Y ~ s(X1) + s(X2), "poisson", data, pval = "chi-square", search = "all")
result$model.opt
```

The `cgam()` function uses `mgcv::gam()` internally and handles smooth
terms (`s()`) in the formula. The Pearson risk invariance principle applies
equally to GAMs.

## Larger example: 5 covariates

In a more realistic setting with 5 candidate covariates and a binomial
response:

```{r five-cov, eval = FALSE}
set.seed(12)
n <- 3000
X1 <- rnorm(n)
X2 <- rnorm(n, X1, 0.5)
X3 <- rnorm(n, 0, 1)
X4 <- rnorm(n, X2, 0.5)
Y  <- rbinom(n, 1, exp(0.8 * X2 - 0.9 * X3) / (1 + exp(0.8 * X2 - 0.9 * X3)))
flip <- rbinom(n, 1, 0.1)
X5 <- (1 - flip) * Y + flip * (1 - Y) + rnorm(n, 0, 0.3)
dat <- data.frame(X1, X2, X3, X4, X5, Y)

# Exhaustive search (evaluates all 2^5 - 1 = 31 subsets)
set.seed(1)
mod_all <- cglm(Y ~ X1 + X2 + X3 + X4 + X5, "binomial", dat,
                pval = "bootstrap", search = "all")
mod_all$model.opt
#> [1] "Y ~ X2 + X3"

# Stepwise search (much faster)
set.seed(1)
mod_step <- cglm(Y ~ X1 + X2 + X3 + X4 + X5, "binomial", dat,
                 pval = "bootstrap", search = "stepwise")
mod_step$model.opt
#> [1] "Y ~ X2 + X3"
```

Both search strategies correctly identify `X2` and `X3` as the causal
parents of `Y`, despite the presence of correlated (X1, X4) and
confounded (X5) variables.

## Summary of recommendations

| Scenario | `family` | `pval` | `search` |
|----------|----------|--------|----------|
| Count data, few covariates (p < 15) | `"poisson"` | `"chi-square"` | `"all"` |
| Count data, many covariates | `"poisson"` | `"chi-square"` | `"stepwise"` |
| Binary data, few covariates | `"binomial"` | `"bootstrap"` | `"all"` |
| Binary data, many covariates | `"binomial"` | `"bootstrap"` | `"stepwise"` |
| Nonlinear effects | Use `cgam()` with `s()` terms | Same as above | Same as above |

## Reference

Polinelli, A., V. Vinciotti and E.C. Wit. (2026). "Causal generalized
linear models via Pearson risk invariance." *Journal of Causal Inference*.
